A statistical framework for analysing whole-genome bisulfite-sequencing data

### 2015-09-15 Rmarkdown/comethylation_review.Rmd
Closes #19

\begin{chapabstract}
Co-methylation is the dependence structure of DNA methylation data. This chapter critically reviews different definitions of co-methylation and puts them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}. It highlights various shortcomings of existing methods for quantifying co-methylation, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}. We also review how methods to detect differential methylation have sought to account for, and exploit, co-methylation.
\end{chapabstract}

@@

There are also reasons for studying co-methylation that are less to do with improving our biological understanding and more to do with improving statistical techniques for analysing DNA methylation data. If co-methylation can be better estimated and understood, then statistical techniques can be developed that explicitly account for and even exploit these dependencies to create more powerful and efficient analysis methods.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/Lyko_Figure_S7.pdf}
\caption[Co-methylation figure from \citet{Lyko:2010dr}]{Periodicity of methylation patterns. (A) Autocorrelation of CpG methylation status over 1 kb. (B) Autocorrelation over 100 bp. Figures A and B show that the correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away. CpGs within a neighborhood of 2 bp to 100 bp are thus more likely to share the same methylation status than more distant CpGs. (C) Fourrier transform of autocorrelation showing a clear periodicity peak at 33 cycles per 100 bp (every 3 bp). (D) Distribution of codon position of mCs, and distribution of methylation level depending on the position. These two panels indicate that the distance between methylated CpGs is often a multiple of three and that the methylated cytosine corresponds most frequently to the first nucleotide of an arginine codon. A reproduction of Supplementary Figure 7 from \citet{Lyko:2010dr} under the Creative Common Attribution (CCBY) licence.}
\label{fig:Lyko}
\end{figure}

@@

In order to exploit between-sample co-methylation in this way, we also need to know the size of these co-methylation blocks, that is, the length of the regions over which between-sample co-methylation is sufficiently strong. The longer these blocks are, the more sparsely we can afford to sample individual methylomes.

@@

\section{Exploiting co-methylation in downstream analyses}
\label{sec:exploiting_cometh}

The variety of definitions, rigour, and reproducibility of the published analyses of co-methylation, not to mention the lack of software implementing the proposed methods, has made it difficult to properly account for co-methylatoin in downstream analyses. Ideally, co-methylation would not only be _accounted for_ in downstream analyses but _exploited_ to leverage the power of the correlation structure of DNA methylation data. In fact, some authors have already sought to do this.

For example, the R/Bioconductor packages `bsseq` \citep{Hansen:2011gu, Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} both cite \citet{Eckhardt:2006gh} as providing evidence of co-methylation, which can be exploited by spatial smoothing of the $\beta$-values prior to testing for differential methylation. Specifically, by assuming that CpGs close to one another have similar methylation levels, the smoothed $\beta$-values are, on average, more accurate than the raw $\beta$-values. Both `bsseq` and `BiSeq` smooth $\beta$-values on a per-sample basis.

@@

These methods exploiting correlations amongst the test statistics or their P-values are not directly based on co-methylation estimates. However, what drives the correlations amongst the test statistics is a complex mixture of within-sample and between-sample co-methylation.

@@

It is clearly inappropriate to treat methylation at nearby loci as independent and methods have been proposed to account for or exploit this in various ways.  These methods would be better informed by a more rigorous and reproducible analysis of co-methylation, which I give in Chapter \ref{chap:co-methylation}.

>>>>

\begin{chapabstract}
Co-methylation is the dependence structure of DNA methylation data. This chapter critically reviews different definitions of co-methylation and puts them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}. It highlights various shortcomings of existing methods for quantifying co-methylation, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}. We also review how methods to detect differential methylation have sought to account for, and leverage, co-methylation.
\end{chapabstract}

@@

There are also reasons for studying co-methylation that are less to do with improving our biological understanding and more to do with improving statistical techniques for analysing DNA methylation data. If co-methylation can be better estimated and understood, then statistical techniques can be developed that explicitly account for and leverage these dependencies to create more powerful and efficient analysis methods.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/Lyko_Figure_S7.pdf}
\caption[Co-methylation figure from \citet{Lyko:2010dr}]{Periodicity of methylation patterns. (A) Autocorrelation of CpG methylation status over 1 kb. (B) Autocorrelation over 100 bp. Figures A and B show that the correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away. CpGs within a neighborhood of 2 bp to 100 bp are thus more likely to share the same methylation status than more distant CpGs. (C) Fourrier [sic] transform of autocorrelation showing a clear periodicity peak at 33 cycles per 100 bp (every 3 bp). (D) Distribution of codon position of mCs, and distribution of methylation level depending on the position. These two panels indicate that the distance between methylated CpGs is often a multiple of three and that the methylated cytosine corresponds most frequently to the first nucleotide of an arginine codon. A reproduction of Supplementary Figure 7 from \citet{Lyko:2010dr} under the Creative Common Attribution (CCBY) licence.}
\label{fig:Lyko}
\end{figure}

@@

In order to leverage between-sample co-methylation in this way, we also need to know the size of these co-methylation blocks, that is, the length of the regions over which between-sample co-methylation is sufficiently strong. The longer these blocks are, the more sparsely we can afford to sample individual methylomes.

@@

\section{Leveraging co-methylation in downstream analyses}
\label{sec:levaraging_cometh}

The variety of definitions, rigour, and reproducibility of the published analyses of co-methylation, not to mention the lack of software implementing the proposed methods, has made it difficult to properly account for co-methylatoin in downstream analyses. Ideally, co-methylation would not only be _accounted for_, but _leveraged_, in downstream analyses of DNA methylation data. In fact, some authors have already sought to do this.

For example, the R/Bioconductor packages `bsseq` \citep{Hansen:2011gu, Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} both cite \citet{Eckhardt:2006gh} as providing evidence of co-methylation, which can be leveraged by spatial smoothing of the $\beta$-values prior to testing for differential methylation. Specifically, by assuming that CpGs close to one another have similar methylation levels, the smoothed $\beta$-values are, on average, more accurate than the raw $\beta$-values. Both `bsseq` and `BiSeq` smooth $\beta$-values on a per-sample basis.

@@

These methods leveraging correlations amongst the test statistics or their P-values are not directly based on co-methylation estimates. However, what drives the correlations amongst the test statistics is a complex mixture of within-sample and between-sample co-methylation.

@@

It is clearly inappropriate to treat methylation at nearby loci as independent and methods have been proposed to account for or leverage this in various ways.  These methods would be better informed by a more rigorous and reproducible analysis of co-methylation, which I give in Chapter \ref{chap:co-methylation}.

### 2015-09-15 Rmarkdown/methsim.Rmd
Closes #19

1. Realism: The simulated data must be 'similar' to the real data. While this is obvious, it is also often hard to pin down or agree upon what constitutes 'similar enough'.
2. Cost: It should be fast and cheap to simulate data. The most common use of simulation models in applied statistics is to repeatedly generate datasets under a range of parameter settings. This requires that each simulation is fast and computationally cheap, otherwise it will be prohibitive to explore the full space of scenarios. An exception to this rule may occur when the simulation model is used to test a proposed mechanistic or stochastic model of a phenomenon, such as in studies of molecular dynamics. Even then, however, the cost of a simulation should be less than the cost of performing the equivalent experiments, otherwise the simulation is generally not worth the effort.
3. Usability: There __must__ be a software implementation. Simulation models exist to be simulated from; a simulation model without an implementation is next to useless. The implementation should give the user easy access to the key  parameters and have sensible default settings. The output of the software should be in a standard format or readily convertible to a simple, manipulable format.

@@

\begin{description}
  \item[\emph{I}] The sample is 'pure' and all haplotypes have the same parameters.
  \item[\emph{II}] The sample is a mixture of subpopulations with different parameters. We will consider a sample with four subpopulations with relative frequencies equal to $(0.6, 0.25, 0.1, 0.05)$.
\end{description}

>>>>

1. Realism: The simulated data must be 'similar' to the real data. While this is obvious, it is also often hard to clearly define or agree upon what constitutes 'similar enough'.
2. Cost: It should be fast and cheap to simulate data. The most common use of simulation models in applied statistics is to repeatedly generate datasets under a range of parameter settings. This requires that each simulation is fast and computationally cheap, otherwise it will be prohibitive to explore the full space of scenarios. An exception to this rule may occur when the simulation model is used to test a proposed mechanistic or stochastic model of a phenomenon, such as in studies of molecular dynamics. Even then, however, the cost of a simulation should be less than the cost of performing the equivalent experiments, otherwise the simulation is generally not worth the effort.
3. Usability: There __must__ be a software implementation. Simulation models exist to be simulated from; a simulation model without an implementation is next to useless. The implementation should give the user easy access to the key  parameters and have sensible default settings. The output of the software should be in a standard format or readily convertible to a simple, manipulable format.

@@

\begin{description}
  \item[\emph{I}] The sample is `pure' and all haplotypes have the same parameters.
  \item[\emph{II}] The sample is a mixture of subpopulations with different parameters. We will consider a sample with four subpopulations with relative frequencies equal to $(0.6, 0.25, 0.1, 0.05)$.
\end{description}

### 2015-09-15 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Closes #19

While 1-tuples are the basis of most analyses of bisulfite-sequencing data, they do not always give the complete picture of how DNA methylation is acting in the sample. To gain a clearer picture, we can exploit the fact that many bisulfite-sequencing reads contain multiple methylation loci (_m-tuples_) and that each read is from a single cell[^chimeric_reads]. An example of where this is useful is shown in Figure \ref{fig:incomplete_picture_1-tuples} where we have two regions, each with four methylation loci, that have identical methylation calls at 1-tuples yet very different overall methylation patterns. Further examples of where m-tuples are useful are in studying _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).

>>>>

While 1-tuples are the basis of most analyses of bisulfite-sequencing data, they do not always give the complete picture of how DNA methylation is acting in the sample. To gain a clearer picture, we can leverage the fact that many bisulfite-sequencing reads contain multiple methylation loci (_m-tuples_) and that each read is from a single cell[^chimeric_reads]. An example of where this is useful is shown in Figure \ref{fig:incomplete_picture_1-tuples} where we have two regions, each with four methylation loci, that have identical methylation calls at 1-tuples yet very different overall methylation patterns. Further examples of where m-tuples are useful are in studying _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).

### 2015-09-15 Rmarkdown/wgbs_statistical_framework.Rmd
Closes #19

Both `bsseq` and `BiSeq` use a binomial local likelihood smoother. In each case this smoother is chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | d_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$. The smoothing is 'local' to exploit co-methylation, which is considered a local phenomenon.

@@

All three _Seisenberger_ samples have noticeably less correlated strand-specific $\beta$-values, including the embryonic stem cell, _J1\_1_. However, since these data are from pooled DNA, the source of this reduced correlation is difficult to pin down.

>>>>

Both `bsseq` and `BiSeq` use a binomial local likelihood smoother. In each case this smoother is chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | d_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$. The smoothing is 'local' to leverage co-methylation, which is considered a local phenomenon.

@@

All three _Seisenberger_ samples have noticeably less correlated strand-specific $\beta$-values, including the embryonic stem cell, _J1\_1_. However, since these data are from pooled DNA, the source of this reduced correlation is difficult to identify.
